Bootstrap learning via modular concept discovery

### 2013-01-26 drafts/ijcai2013/ijcai13.tex
Commits.

\title{Getting off the ground: bootstrap learning via modular concept discovery.}

@@

\begin{abstract}
 Suppose a learner is faced with a domain of problems about which she
 knows nearly nothing. She does not know the distribution of problems,
 the space of solutions is not smooth, and the reward signal is
 uninformative, providing perhaps a few bits of information but not
 enough to steer her effectively. How can such a learner ever get off
 the ground? A common intuition is that if the solutions to these
 problems share a common structure, and the learner can solve some
 simple problems by luck or brute force, she should be able to extract
 useful modular concepts from these solutions and, by composing them,
 explore the solution space more efficiently. Here, we formalize this
 intuition, where the solution space is the space of typed functional
 programs and the gained information is stored as a stochastic grammar
 over programs. We propose a two step iterative procedure for
 exploring such spaces: in the first step, the learner explores a
 finite subset of the domain, guided by a stochastic grammar; in the
 second step, the learner compresses the successful solutions from the
 first step to estimate a new stochastic grammar. As we show in a
 symbolic regression and a boolean circuit learning example, iterating
 these two steps quickly propels the learner from a state of solving
 almost no problems to solving a large number of problems by gaining
 abstract knowledge of the structure of the solution space.
\end{abstract}

@@

      \path (and3)  edge node[inner sep=0mm,pos=.4] (t1) {} node {} (l3);
      \path (and3)  edge node[inner sep=0mm,pos=.4] (k1) {} node {} (r3);
      \path (t1) edge[bend right] (k1) ;
    \end{tikzpicture}
    \label{fig:andor}
    \end{subfigure}
  \caption{(a) The typed combinator $S * I$ ($f(x)=x^2$) represented as
    binary tree. (b) The space of all typed combinators represented as an
    infinitely deep AND/OR tree. Any specific combinator is a
    sub-binary tree of this AND/OR tree whose leaves consist of only
    the $f_i$ nodes (i.e., the primitive combinators of the grammar).}

@@

\subsection{A stochastic grammar over programs.}
A stochastic grammar over programs represents a distribution over
expressions in $\mathcal{L}$. We represent our grammar as a
multinomial distribution over the primitives in our library. Let $p_1,
\dots, p_m$ be the probabilities associated with primitive combinators
$c_1, \dots, c_m$. Then we define the probability $p(c_i | t) $ of
using $c_i$ at a node whose type is $t$ to be $p_i$ normalized by the
probability mass of all other primitive combinators that could have
been used at that node. Namely, $p(c_i | t) = \frac{p_i}{\sum_{c_j :
    type(c_j) \text{ unifies with } t} p_j}$ if type($c_i)$ unifies
with $t$, 0 otherwise. Note that the distribution thus defined is not
normalizable if we allow expressions of arbitrary depth, but this is
not a practicle concern since we are using the distribution mainly as
a search heuristic.

\subsection{Best-first enumeration of most probable programs.}
In order to explore the frontier of most promising programs, we need a
procedure that enumerates the $N$ most probable programs. There has
been recent interest in this
problem~\cite{DBLP:conf/sfp/Katayama05}\cite{DBLP:conf/aaip/YakushevJ09}. Our
approach is to formulate the problem as a best-first exploration of an
AND/OR tree~\cite{nilsson1982principles}, which we denote
$\mathcal{G}$. A complete program is any binary subtrees of
$\mathcal{G}$ whose leaves coincide with the leaves of
$\mathcal{G}$. The tree structure is shown in
Figure~\ref{fig:andor}. The root is an OR node of the requisite
program type. It's children are the primitives of that type $f_1,
\dots, f_n$, and one AND node with two children. Each child of an AND
node has the same structure as the root node, with modified types: if
the type of an AND node is $t$, then the type of its left child is
$\tau \rightarrow t$ (where $\tau$ is a fresh type variable). Since
the type of the right child is constrained by the subtree rooted at
the left child, we always expand the left child first and use the
target type of this left subtree to mark the type of the right child.

In accordance with the stochastic grammar over programs, we weight all
edges from OR nodes according to the negative log likelihood of
choosing the associated rule from the stochastic grammar, conditioned
on the requesting type. Expanding an AND node has zero cost. In the
case that the stochastic grammar is ill conditioned, we might have an
infinitely deep solution tree whose cost does not increase as we
unfold it; this can occur, for example, if there are no primitives of
the required type since the log probability of expansion to the next
AND node is 0. To avoid this, we simply declare a cost for
expanding the next AND node when it is the only node available.

@@

Once we have enumerated the best $N$ programs from our distribution,
we will find that some tasks have no solutions and some have one or
more solutions; for those tasks that have solutions, we want to choose
a solution for each task such that the complete set of such solutions
for all the tasks is as compressible as possible.

A natural metric for the compressibility of a set of binary trees is
the number of unique subtrees that this set contains. Thus, we can
attempt to find the assignment of solutions to tasks that minimizes
this value. However, an exact optimization for this problem requires
examining $O(m^k)$ assignments, where $m$ is the maximum number of
solutions for any one task and $k$ is the number of tasks. Since this
is prohibitive, we relax the compressibility metric as follows: we
define the cost of adding each new solution $s_n$ to a partial set of
solutions $s_1, \dots, s_{n-1}$ to be the number of additional unique
subtrees in $s_n$ that are not present in $s_{n-1}$. Our solution thus
becomes approximate and that approximation is order-dependent, but a
depth-first search on the defined search space goes from exponential
in the solution degree to quadratic.

@@

%%(1,0) node[nand port, thick] (nor1) {}
(3, 0)  node[ rotate=0, anchor=east,thick, draw, text width=1cm ] (f) {?} 
(1, 0) node[nand port, thick] (nor) {}
(nor.in 1) node[above left=.25cm] (x) {x}
(nor.in 2) node[below left=.25cm] (y) {y}
(y) -- (1.71,.5cm)
(nor) -- (1.72,-.5cm)
(x) -| (nor.in 1) 
(y) -| (nor.in 2);
\end{circuitikz}
\caption{A schematic that represents the reused component
  \texttt{\small{(B (C B nand) S)}}.}
\label{fig:component}
\end{figure}

>>>>

\title{Bootstrap learning via modular concept discovery}

@@

\begin{abstract}
 Suppose a learner is faced with a domain of problems about which it knows nearly nothing. It does not know the distribution of problems, the space of solutions is not smooth, and the reward signal is uninformative, providing perhaps a few bits of information but not enough to steer it effectively. How can such a learner ever get off the ground? A common intuition is that if the solutions to these problems share a common structure, and the learner can solve some simple problems by brute force, it should be able to extract useful components from these solutions and, by composing them, explore the solution space more efficiently. Here, we formalize this intuition, where the solution space is the space of typed functional programs and the gained information is stored as a stochastic grammar over programs. We propose an iterative procedure for exploring such spaces: in the first step of each iteration, the learner explores a finite subset of the domain, guided by a stochastic grammar; in the second step, the learner compresses the successful solutions from the first step to estimate a new stochastic grammar. We test this procedure on symbolic regression and boolean circuit learning and show that the learner discovers modular concepts for these domains. Whereas the learner is able to solve almost none of the posed problems in the procedure's first iteration, it rapidly becomes able to solve a large number by gaining abstract knowledge of the structure of the solution space.

\end{abstract}

@@

      \path (and3)  edge node[inner sep=0mm,pos=.4] (t1) {} node {} (l3);
      \path (and3)  edge node[inner sep=0mm,pos=.4] (k1) {} node {} (r3);
      \path (t1) edge[bend right] (k1) ;
    \end{tikzpicture}
    \caption{AND/OR expression search space. }
    \label{fig:andor}
    \end{subfigure}
  \caption{(a) The typed combinator $S * I$ ($f(x)=x^2$) represented as
    binary tree. (b) The space of all typed combinators represented as an
    infinitely deep AND/OR tree. Any specific combinator is a
    sub-binary tree of this AND/OR tree whose leaves consist of only
    the $f_i$ nodes (i.e., the primitive combinators of the grammar).}

@@

\subsection{A stochastic grammar over programs.}
\label{sec:stochgrammar}
A stochastic grammar over programs represents a distribution over
expressions in $\mathcal{L}$. We associate with each primitive
combinator $c_i$ in $\mathcal{L}$ a probability $p_i$,
$0~\leq~p_i~\leq~1, \sum_i~p_i~=~1 $.

We define the probability of an expression $e$ consisting of primitive
combinators $c_1, \dots, c_N$ to be $p(e)~=~\prod_{n=1}^N p(c_n |
t(c_n))$ where $t(c_n)$ is the requested type when $c_n$ was used and
\[
p(c_i | t(c_n)) =    \frac{1}{4}\frac{p_i}{\sum_{c_j :
    type(c_j) \text{ unifies with } t(c_n)} p_j}.
\]
That is, $p(c_i | t(c_n)$ is the probability of sampling $c_i$ from
the multinomial distribution defined by the $p_i$'s, conditioned on
the requesting type, and then divided by 4. The constant factor of
$1/4$ is introduced as a simple way to ensure that the sum over the
probabilities of all expressions converges to finite value (so that
the distribution over expressions is proper).\footnote{The probability
  mass $M_d$ of all expressions of depth less than or equal to $d$ can
  be written with the recurrence relation $M_d \leq M_d^2 + M_1$. This
  has an asymptotic fixed point if $x = x^2 + M_1$ has a solution,
  which is true if $M_1 \leq 1/4$.}

To estimate the values $p_i$ from set of expressions, we simply divide
the number of times primitive combinator $c_i$ was used by the number
of times it could have been used given the requesting types in the
expression trees. Note that this is not a maximum likelihood estimator
for this distribution; finding that would require solving a rather
complex optimization problem because of each choice is conditioned on
the requesting type. We choose our estimator merely for convenience,
and find that our empirical results justify this simplicity. It is a
matter for future research to determine to what extent a more accurate
estimator would improve performance.

 \subsection{Best-first enumeration of most probable programs.}
In order to explore the frontier of most promising programs, we need a
procedure that enumerates the $N$ most probable programs. There has
been recent interest in this problem, most of it focused on
enumerating the shortest program satisfying some
criterion~\cite{DBLP:conf/sfp/Katayama05} \cite{DBLP:conf/aaip/YakushevJ09}. Our
approach is to formulate the problem as a best-first exploration of an
AND/OR tree~\cite{nilsson1982principles}, which we denote
$\mathcal{G}$. A complete program is any binary subtrees of
$\mathcal{G}$ whose leaves coincide with the leaves of
$\mathcal{G}$. The tree structure is shown in
Figure~\ref{fig:andor}. The root is an OR node of the requisite
program type. It's children are the primitives of that type $f_1,
\dots, f_n$, and one AND node with two children. Each child of an AND
node has the same structure as the root node, with modified types: if
the type of an AND node is $t$, then the type of its left child is
$\tau \rightarrow t$ (where $\tau$ is a fresh type variable). Since
the type of the right child is constrained by the subtree rooted at
the left child, we always expand the left child first and use the
target type of this left subtree to mark the type of the right child.

Recall that a valid partial solution is a subtree of $\mathcal{G}$
satisfying the following properties: its root is the root
$\mathcal{G}$ and any OR node has at most one child. A complete
solution is a partial solution whose leaves are leaves of
$\mathcal{G}$, whose root is the root of $\mathcal{G}$, in which each
OR node has exactly one child, and in which each AND node has exactly
two children.

The value of a partial solution $G'$ is calculated by summing the
value of its leaves. If a leaf of $G'$ is a leaf of $G$ then either it
is an AND node, to which we assign a value of 0, or it corresponds to
some primitive combinator $c_i$ in the library and is the child of
some typed OR node with type $t$. We assign its value of $- \log{p(c_i
  | t)}$, according to the stochastic grammar specified in
Section~\ref{sec:stochgrammar}. If a leaf $v$ of $G'$ is an OR node,
then the value of any extension of it to a complete solution is at
most $-log(1/4)$. Thus, we have defined a value for partial solutions
that upper bounds the value of any of its extensions.

Our best-first policy, is to simply expand the next partial solution
with the highest value. If a partial solution has multiple leaves that
are not primitive combinators, we expand the left most one first,
since it is the choice of left combinator that defines the type of its
right combinator in an application pair. That is, the parent node
applies its constraint on valid solutions via the left child, not the
right child. We want to apply constraints as early as possible, so
expanding left children first prunes more partial solutions.

\textbf{****NEED COMPLEXITY ANALYSIS HERE*******}

@@

We say that a task is \emph{hit} if it has a solution in the
frontier. Having enumerated the frontier, we want to assign solutions
to hit tasks such that this set is maximally compressible. This will
promote the usage of reusable modular subtrees.

A natural metric for the compressibility of a set of binary trees is
the number of unique subtrees that this set contains. As an example of
this, suppose we had three tasks whose solutions were integers 16, 25,
and 36, and suppose our frontier contained the six expressions:

\begin{figure}[h]
    \begin{tikzpicture}[scale=.9,
        transform shape, level/.style={sibling distance=10mm/#1, font=\scriptsize}
        ]
      \node [font=\scriptsize] (top) {}
      child {node [above=.3cm, text width=2cm] (a) {$(Int \rightarrow Int) \rightarrow Int 
          \rightarrow Int$}
        child {node [left=.5cm, below=1cm, text width=3cm] 
          {$S : \\(t_2 \rightarrow t_1 \rightarrow t_0)  \rightarrow (t_2 \rightarrow t_1) 
            \rightarrow \\ t_2 \rightarrow t_0$}}
        child {node [right=0cm, text width=2cm] {$*: Int \rightarrow Int \rightarrow Int$}}
      }
      child {node [below right=.3cm and 1cm] (b) {$Int \rightarrow Int$}
        child {node {$I: t_0 \rightarrow t_0 \rightarrow t_0$}}
      };
    \end{tikzpicture}
\end{figure}

Let $|e|$ be the
number of unique binary trees in expression $e$.

we will find that some tasks have no solutions
and some have one or more solutions; for those tasks that have
solutions, we want to choose a solution for each task such that the
complete set of such solutions for all solved tasks is as compressible
as possible.

Thus, we can attempt
to find the assignment of solutions to tasks that minimizes this
value. However, an exact optimization for this problem requires
examining $O(m^k)$ assignments, where $m$ is the maximum number of
solutions for any one task and $k$ is the number of tasks. Since this
is prohibitive, we relax the compressibility metric as follows: we
define the cost of adding each new solution $s_n$ to a partial set of
solutions $s_1, \dots, s_{n-1}$ to be the number of additional unique
subtrees in $s_n$ that are not present in $s_{n-1}$. Our solution thus
becomes approximate and that approximation is order-dependent, but a
depth-first search on the defined search space goes from exponential
in the solution degree to quadratic.

@@

%%(1,0) node[nand port, thick] (nor1) {}
(3, 0)  node[ anchor=east, text align=center, thick, draw, text width=1cm, text height=.25cm, align=center] (f) {?} 
node [right of=f] (out) {}
(1, 0) node[nand port, thick] (nor) {}
(nor.in 1) node[above left=.25cm] (x) {x}
(nor.in 2) node[below left=.25cm] (y) {y}
(y) -| (f)
(nor) -- (f)
(f) -| (out)
(x) -| (nor.in 1) 
(y) -| (nor.in 2);
\end{circuitikz}
\caption{A schematic that represents the reused component
  \texttt{\small{(B (C B nand) S)}}.}
\label{fig:component}
\end{figure}
